PhD Proposal

### 2022-04-25 abstract/abstract.tex
continuo abstract

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=20mm, bottom=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

@@

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE,RO]{\slshape \rightmark}
\fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}

@@

proposed its use together with the so-called FM-index \cite{fmindex}.\\
In recent years, in the field of bioinformatics, there is a change of
interest. While until a few years ago the research was focused on the study of
the \textit{genome}, we are now starting to deepen the topic of
\textit{pangenome}. Currently, the need to take into

@@

In this context, various algorithms and various data structures have been
implemented in order to study the so called \textit{haplotypes} and
\textit{genotypes}. Briefly we could define \textit{haplotypes} as a combination
\textit{genotype} as the complete set of genes contained in the DNA. From a
biological point of view it is indeed interesting to note that each individual
shares about 99\% of the genetic code, while the remaining 1\% differs mostly
with \textbf{Single Nucleotide Polymorphisms (\textit{SNPs})}, i.e. variations
of a single nucleotide in a precise \textit{locus} of the DNA\\ 

@@

\textit{pangenome graph}, replacing the old representation of the single genome
by a single sequence. \\

@@

particular data structure it was possible both to study possible matches inside
a haplotype panel and to search for set maximal matches between an external
haplotype and a panel, as well as the set maximal matches inside the panel
itself. Furthermore, for example, variants have also been studied for the 
management of the multiallelic case \cite{mpbwt}. \\

@@

\textbf{run-length version of BWT (\textit{RLBWT})} with the ultimate intention
of indexing and using multiple genomes as a reference \cite{moni}. Together with
this data structure the authors proposed the concept of \textit{matching
  statistics} in order to efficiently compute the matches between a pattern and
a text. A recent improvement, regarding the \textit{RLBWT}, has been made
through the implementation of \textit{PHONI} \cite{phoni}, where where the
so-called \textit{longest-common-extension (LCE) queries} are used. \\

@@

authors of MONI in order to create a run length encoded variant of the PBWT. In
this work I used practically all concepts applied to the ``classic''
\textit{BWT} in \textit{MONI} and \textit{PHONI}. \\
During my PhD I will focus my studies on the development of new algorithms in
various topics such as variants of the \textit{PBWT} (and any related uses),
\textit{haplotyping/genotyping} or solutions concerning the study of
\textit{SNPs}.It is also my intention to deepen the more experimental themes
relating to pattern matching, in detail the new developments on BWT and new
indexing structures, as well as the theme of \textit{succinct data structures},
with particular attention to the use of \textit{bitvectors}. 

@@

methods etc that will be the core of my studies during my PhD. 
\subsection*{BWT}
The \textbf{Burrows-Wheeler Transform (\textit{BWT})} \cite{bwt} was introduced

@@

text $T$, \$-terminated, such that $|T|=n$, we can define the $BWT_T$, denoting
with $SA_T$ the \textit{suffix array} of $T$, as:

@@

the number of occurrences of characters lexicographically than the one given as
argument in $T$. The second one is the $Occ$ function,
$Occ:\Sigma\times[1,n]\to[1,n]$, has as arguments a character $\sigma\in\Sigma$

@@

$BWT_T[1,i]$ (\textbf{CONTROLLARE E RISCRIVERE MEGLIO}).\\

@@

\subsection*{Bitvectors}
\textit{Bitvectors} are ones of the most important data structure when
mentioning succinct data structures. \\

@@

operations possible on Boolean arrays, such as \textit{random access} in
\textit{constant time} (\textbf{DA VERIFICARE}). More in detail the rank
function allows you to calculate how many values of 1 are up to a certain
index. Instead the select function allows to obtain the index of any one present
in the bitvector. Formally, given a bitvector $B$, such that $|B|=n$, and given

@@

constant time.  In more practical terms there are several implementations of the
same within \textbf{SDSL (\textit{Succint Data Structures Library})}
\cite{sdsl}, one of the most important C++ library used in bioinformatics. As

@@

operations varies (usually only one of the two is in constant time) and the
amount of additional bits needed.\\

@@

\subsection*{RLBWT}
Speaking of pangenome, linear indexing via FM-index is no longer the best

@@

text, $n$, to being linear over the number of runs, $r$. \\

@@

al. \cite{rindex}, was called \textbf{r-index} and corresponds to the
\textit{RLBWT} and a SA sampling at the begin and at the end of every run. The

@@

the text) cd\cite{lce} and \textbf{straight-line programs (\textit{SLP})}
\cite{slp} (for \textit{random access} and \textit{lce queries} over the
grammar-compressed text) in \textit{PHONI}. Both these solutions are used for
the computation of the \textbf{matching statistics (\textit{MS})}. Given a
pattern $P$ and a text $T$ the \textit{MS} of $P$ in respect to $P$ is an array

@@

$T[M[i].pos:M[i].pos+M[i].len-1=P[i:i+M[i].len]-1]$ and $P[i:i+M[i].len]$ does

@@

$P[i-1:i+l-1]$ nor $P[i:i+l]$ are. At the end, using a particular function
called $\varphi$ (and $\varphi^{-1}$), based on the use of the \textbf{inverse
  suffix array (\textit{ISA})} as well, it was possible to start from the
starting positon of a match extracted 
by \textit{MS} and find all the starting positions of all the copies of $P$ in
$T$. Infact, formally, given a starting position $p$, we have

@@

\subsection*{PBWT}
Based on the theories of BWT Durbin, in 2014, he devised the positional
\textbf{Burrows–Wheeler transform (\textit{PBWT}}) \cite{pbwt}, in order to
solve the problem of pattern matching on panels (matrices) of haplotypes,
denoted by $X$. In
detail we are talking about a panel with $M$ haplotypes and $N$ biallelic
sites. This data structure is based on a reversed-prefix ordering at each column
$k$ that produces two different multidimensional arrays. The first one is called
\textbf{prefix array}, denoted by $a$, and, for each column $k$, contains, for

@@

could be denoted by $y_i^k$. The second bidimensional array is
called \textbf{divergence array}, denoted by $d$, and indicates the index of the

@@

$y_i^k[h,k)=y_{i-1}^k[h,k)$. (\textbf{RISCRIVERE MEGLIO})\\

@@

in linear time, all set-maximal matches (which we could also call ``MEMs'') from
a new sequence $z$ to $X$ etc$\ldots$ 

@@

\textit{row} of the panel, regarding thresholds we use the index inside a run in
a column where we have the minimum value of divergence array and, regarding the
\textit{SLP}, we ``stretch'' the reverse panel (from the right to the left) in
in order to make \textit{LCE queries} possible. Instead, regarding the $\varphi$
function, I have implemented a new simple data structure to identify which row
is above and which row is below each row in the panel permutated via the
prefix array and stored in a run-length compressed way. (\textbf{RISCRIVERE
  MEGLIO ED EVENTUALMENTE APPROFONDIRE}).\\

@@

the case of real data. First of all it will be interesting to implement a new

@@

current use of bitvectors to manage the \textit{LF-mapping} also in that
case. On the other hand, it will be more complicated to manage the missing
data. This problem is known to be \textit{hard} so I should probably deal with
parametric algorithms or approximate algorithms, based on researches already
developed in \textit{BIAS}.

@@

>>>>

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=20mm, right=20mm, top=20mm, bottom=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

@@

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
% \pagestyle{fancy}
% \fancyhead[LE,RO]{\slshape \rightmark}
% \fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}

@@

proposed its use together with indexing via \textit{FM-index}.\\
Furthermore, in recent years, in the field of bioinformatics, there has been a
change of interest. While until a few years ago the research was focused, as
anticipated, on the study of a \textbf{linear sequence of a genome}, the
researchers have begun to deepen the topic of the \textbf{pangenome}, which
term was 
introduced by Tettelin in 2005. Infact the need to take into 

@@

@@

\textbf{pangenome graph}. \\
From a biological point of view it is indeed interesting to note that, as
pointed out by the study of \textit{1000 Genome Project}, over 88 million
variants, 84.7 million are \textbf{Single Nucleotide Polymorphisms
  (\textit{SNPs})}, 3.6 million are \textbf{short insertions/deletions
  (\textit{indel})} and 60000 \textbf{structural variants}, involving more than
50 nucleotides. All these variants are now a limit to the traditional use of a
\textit{linear sequence of a genome}.\\
Against this background, various algorithms and various data structures have
been implemented in order to study \textit{haplotypes} and
\textit{genotypes}, for example to study the \textbf{genotyping variants
  problem}. Briefly we could define \textit{haplotypes} as a combination 
\textit{genotype} as the complete set of genes contained in the DNA.\\ 

@@

particular data structure (which will be described below) it is possible to
study efficiently a collection of haplotypes but only in the bi-allelic cases,
in the original implementation. Furthermore, for example, variants 
of the \textit{PBWT} have also been studied for handling the multiallelic
case. \\ 

@@

\textbf{run-length encoded version of BWT (\textit{RLBWT})} with the ultimate
intention of indexing and using multiple genomes as a reference
\cite{moni}. Together with this data structure the authors proposed the concept
of \textbf{matching statistics (\textit{MS}} in order to efficiently compute
the matches between a pattern and a text. A recent improvement, regarding the
\textit{RLBWT}, has been made through the implementation of \textit{PHONI}
\cite{phoni}, where the \textbf{longest-common-extension (\textit{LCE}) queries}
are used to compute the \textit{len} of \textit{MS} in a single sweep over the
pattern. \\ 

@@

authors of \textit{MONI} and \textit{PHONI} in order to create a run length
encoded variant of the \textit{PBWT}, the \textbf{RLPBWT}, using and adapting
the various theories developed for the \textit{RLBWT}. In this context my PhD
will be focused on the development of new algorithms in
various topics related to open problems in the study of the \textit{pangenome
  graph}, of the \textit{haplotyping/genotyping} topics and \textit{genome
  variants} problems. It is also my intention to deepen the
more experimental themes 
relating to \textit{pattern matching}, in detail the new developments on
\textit{BWT} and new indexing structures, as well as the theme of
\textit{succinct data structures}, with particular attention to the use of
\textit{bitvectors}.

@@

methods etc $\ldots$ that will be the core of my studies during my PhD. 
\subsubsection*{BWT}
The \textbf{Burrows-Wheeler Transform (\textit{BWT})} was introduced

@@

text $T$, \$-terminated, such that $|T|=n$, we can define, denoting
with $SA_T$ the \textit{suffix array} of $T$, the $BWT_T$ as:

@@

the number of occurrences of characters lexicographically minor than the one
given as argument in $T$. The second one is the $Occ$ function,
$Occ:\Sigma\times[1,n]\to[1,n]$, which has as arguments a character
$\sigma\in\Sigma$ 

@@

$BWT_T[1,i]$.\\

@@

\subsubsection*{Bitvectors}
\textbf{Bitvectors} are ones of the most important data structure when
mentioning \textit{succinct data structures}. \\

@@

operations possible on boolean arrays, such as \textit{random access} in
\textit{constant time}. More in detail the \textit{rank
function} allows to calculate how many values of one are up to a certain
index. Instead the \textit{select function} allows to obtain the index of any
one present 
in the \textit{bitvector}. Formally, given a bitvector $B$, such that $|B|=n$,
and given 

@@

\textit{constant time}. In more practical terms there are several
implementations of the 
same within \textbf{SDSL (\textit{Succint Data Structures Library})}, one of the
most important C++ library used in bioinformatics. As 

@@

operations varies (usually only one of the two is in constant time) as well as
the amount of additional bits needed.\\

@@

\subsubsection*{RLBWT}
Speaking of \textit{pangenome}, linear indexing via FM-index is no longer the
best 

@@

text, $n$, to being linear over the number of runs, $r$, so sub-linear over the
length of the text. \\ 

@@

al., was called \textbf{r-index} and corresponds to the
\textit{RLBWT} and a \textit{suffix array sampling} at the begin and at the end
of every run. The 

@@

the text) in \textit{PHONI}. Both solutions also make use of
\textbf{straight-line programs (\textit{SLP})}, for \textit{random 
  access} in \textit{MONI} and for \textit{lce queries} in \textit{PHONI}.
The purpose of the two projects is computation of the \textbf{matching
  statistics (\textit{MS})}. Given a 
pattern $P$ and a text $T$, the \textit{MS} of $P$ in respect to $T$ is an
array 

@@

$T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and $P[i:i+M[i].len]$ does

@@

$P[i-1:i+l-1]$ nor $P[i:i+l]$ are. Furthermore, using a particular function
called $\varphi$ (and and its ``inverse'' $\varphi^{-1}$), based on the use of
the \textbf{inverse 
  suffix array (\textit{ISA})} as well, it was possible to find all the starting
positions of all the copies of $P$ in $T$ from the starting positon of a match
extracted by \textit{MS}. More formally, given a starting position $p$,

@@

\subsubsection*{PBWT}
Based on the theories of the \textit{BWT} Durbin, in 2014, devised the
\textbf{positional Burrows–Wheeler transform (\textit{PBWT})}, in order to
solve the problem of pattern matching on panels (matrices) of haplotype. In
detail he defined a panel $X$ with $M$ haplotypes and $N$ biallelic
sites. This data structure is based on a \textit{reversed-prefix ordering} at
each column 
$k$ that produces two different multidimensional arrays. The first one is the
set of the
\textbf{prefix arrays}, denoted by $a$, which, for each column $k$, contains, for 

@@

could be denoted by $y_i^k$. The second bidimensional array is the set of the
\textbf{divergence arrays}, denoted by $d$, which indicates the index of the

@@

$y_i^k[h,k)=y_{i-1}^k[h,k)$. \\

@@

in linear time and all set-maximal matches (which we could also call ``MEMs'')
from a new sequence $z$ to $X$ in $\mathcal{O}(M^2N)$.\\
Despite the fact that \textit{PBWT} has been poorly regarded in the scientific
community 
in the early years since its development, there has been a growth in research
based on it, both in terms of variant design, such as the already mentioned
multiallelic version or, for example, the \textbf{dynamic PBWT
  (\textit{d-PBWT}}), and in terms of its use to study haplotype panels, for
imputations etc$\ldots$
\subsubsection*{Studies on haplotypes}
For the sake of completeness, it is necessary to cite some examples of studies
on haplotypes. \\
A first example is \textbf{Ranbow}, proposed by Moeinzadeh et al. in 2020. The
aim of this project was the haplotype reconstruction of polyploid genome from
short read sequencing data, studying also the multi-allelic case. In that case
the \textit{PBWT} was not used but the same could allow a better efficiency of
the algorithm itself.\\
Another example could be found in the paper of Alanko et al., published in
2021, where the authors used the \textit{PBWT} to look for \textit{maximal
  perfect haplotype block}, within an haplotypes panel. These types of
algorithms are essential for the identification of genomic regions that show
signatures of natural selection.\\
Another interesting paper is the one by Wlliams and Mumey, published in 2020,
who also studied \textit{maximal perfect haplotype blocks}, but with the
addition of \textit{missing data}, handled with the help of wildcards, using the
\textit{PBWT}. \\ 
Both these last
two examples are limited in the bi-allelic case but they show the potential of
\textit{PBWT} in the study of haplotypes. However, it is important to note that
in both cases the authors not only make use of \textit{PBWT} but also other data
structures are cited to achieve the same results. 

@@

\textit{row} of the panel. More formally, given a panel $X$ (where every row is
defined by $X_j$) and a pattern $P$,
$|P|=X_{width}=n$, we define \textit{MS} as an array of length $n$ such that,
for each position $1\leq i\leq n$, $X_{MS[i].row}[i-MS[i].len+1:
i]=P[i-MS[i].len+1:i]$ and $P[i-MS[i].len:i]$ is not a suffix of
$X_1[1:i],\ldots,X_{X_{height}}[1:i]$. Talking about thresholds they are defined
by the index of the minimum \textit{LCP value} inside a run, considering also
the \textit{LCP value} of the head of the next run, if exists. Regarding
the \textit{SLP}, to get it, we stretch the reverse panel (from the right to
the left) in 
in order to make \textit{LCE queries} possible, having that \textit{LCE queries}
are made, between two rows, from the right to the left. Instead, for the
$\varphi$  function and its inverse, I have implemented a new simple data
structure to identify which row is above and which row is below each row in the
\textit{RLPBWT matrix}, that is the panel already permuted via the prefix array
and run-length encoded.\\

@@

real cases. First of all it will be interesting to implement a new

@@

current use of bitvectors to manage the \textit{LF-mapping} also with more than
two alleles. On the other hand, it will be more complicated to manage the
missing data. This problem is known to be \textit{hard} so I should probably
deal with parametric algorithms or approximate algorithms, based on researches
already developed in \textit{BIAS}.\\
Summarizing some of the research goals during my \textit{Phd} will be:
\begin{enumerate}
  \item *
  \item *
  \item *
\end{enumerate}
From a more technical point of view, during both my bachelor and master
thesis projects, I mainly focused on the use of the \textbf{C++ programming 
  language},  due to the availability of efficient libraries, as the already
cited \textit{SDSL}, that have now become standard in the bioinformatics
field. In addition to \textit{C++}, however, I had the opportunity to deepen
\textbf{Python}, with libraries such as \textit{biopython}, and \textbf{Rust},
with recently developed libraries such as \textit{bio-rust} (even if not
yet complete from an algorithmic point of view). \\
Another point of interest is the analysis and development of efficient
algorithms based on parallel computing, also on GPU, algorithms that are
increasingly in use in both \textit{bioinformatics} and \textit{systems
  biology}. \\ 
To conclude this proposal I also point out the intention to maintain contacts
with various researchers, initially obtained during the work on the project of
the master thesis, including Christina Boucher (University of Florida), Travis
Gagie (Dalhousie University) etc$\ldots$ in order to make my PhD program more
complete. 

@@

\newpage
\noindent
\begin{shaded}
  \noindent
  \textit{Essendo una prima bozza probabilmente è tutta da riscrivere,
  soprattutto in luce del ``come vada scritta''.}\\
  \textbf{Note}:
  \begin{itemize}
    \item il font dovrebbe essere corretto secondo le richieste. Non ho visto
    specifiche relative ai margini del file
    \item nell'introduzione mancano alcuni concetti relativi allo
    studio dei grafi nel caso pangenomico. Una volta chiariti gli scopri precisi
    del progetto di ricerca bisogna correggere (oltre che aggiungere eventuali
    aspetti non considerati e toglierne altri)
    \item nell'introduzione devo capire come ordinare meglio i concetti
    \item nell'introduzione forse servono esempi di casi d'uso relativi alle
    varie problematiche. Probabilmente è necessaria una parte più discorsiva
    relativa anche all'importanza biologica di questo genere di studi
    \item la conclusione dell'introduzione, dove si parla di cosa verrà
    eventualmente approfondito durante il PhD, è praticamente un
    \textit{placeholder}  
    \item l'incipit allo stato dell'arte è sicuramente da modificare
    \item la struttura dell'intera sezione riguardante lo stato dell'arte penso
    debba essere rivista in base a standard per la stesura della proposal che
    non conosco
    \item nello stato dell'arte la parte relativa alla BWT è commentata in
    quanto, pur avendola scritta, non la ritengo essenziale (soprattutto avendo
    a che fare con un limite di 4 pagine)
    \item nello stato dell'arte la sezione relativa ad alcuni esempi di studio
    sugli aplotipi deve essere profondamente rivista in luce soprattutto dei
    research goals. Un esempio potrebbe essere la citazione, ad esempio, di
    HapCol, qualora i goals si spostassero un po' dalla PBWT e dai suoi usi più
    immediati 
    \item per ignoranza mia su come vada scritto questo tipo di documento non
    sono sicuro che alcuni dettagli formali nei vari passaggi dello stato
    dell'arte siano sensati da mettere. Stesso discorso vale nell'incipit dei
    research goals
    \item la citazione di quanto svolto per la tesi magistrale non so se sia
    necessario, e, qualora lo fosse, se la collocazione all'inizio dei research
    goals sia corretta, ne tantomeno se sia troppo riassunto/esteso
    \item mancano le specifiche dei fini della ricerca, da chiarire coi docenti
    (e da cui dipendono diversi punti della proposal)
    \item la conclusione del research goals non so se sia necessaria
    \item per quanto riguarda la scelta delle reference nel file \texttt{.bib}
    ne sono 
    presenti diverse. In ogni caso, per la prima selezione fatta, ritengo le
    prime due necessarie mentre la terza e la quarta sono per diversi punti di
    vista superflue. Avendo un bound di cinque citazioni c'è comunque spazio di
    manovra
  \end{itemize}
\end{shaded}
